---
title: "GAM_nongaussian"
output: html_notebook
---

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)

source('gam_functions.R')
source('data_functions.R')
```

# Portal Plot Flip using gam()

This notebook tries to apply factor-smooth approach outlined by Gavin 
Simpson in his [blog post](https://www.fromthebottomoftheheap.net/2017/10/10/difference-splines-i/)
and combine it with the trend-season appraoch outlined in his other series of
posts: [modelling seasonal data with GAMS](https://www.fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/),
[are some seasons warming more than others?](https://www.fromthebottomoftheheap.net/2015/11/23/are-some-seasons-warming-more-than-others/)

Here we go:

## Load Data

Data is coming straight from the repo. This code is extracting the dipo 
abundances from the straight controls (CC), and the opened krat (EC) and rodent 
(XC) exclosures. This data runs from period 415 (3/13/2013) to present

```{r}
data = get_data()
dipoN = make_N_data(species = 'Dipos', data)
dipoN$month = as.numeric(format.Date(dipoN$censusdate, "%m"))
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
```

## Basic Season-trend model with factors

Here's a grap hof the basic patterns in the data by treatment type. We see
the pure control plots (green) are relative stable through time (dip in most 
recent months) and the increase of the kangaroo rats on the two opened plots, 
with the number of kangaroo rats appearing higher on the opened rodent 
exclosures (blue) than the kangaroo rat exclosures (red)


```{r}
ggplot(dipoN, aes(x = censusdate, y = n, colour = treatment)) +
    geom_point() +
    geom_smooth(method = 'loess', se = FALSE) +
    scale_colour_brewer(type = 'qual', palette = 'Dark2') +
    theme(legend.position = 'top')
```

## GAM model

Currently, this model allow for a trend-season interaction (changes in the
seasonal signal through time). This may be overkill. We use a tweedie because
when we were fitting the treatment separately, the tweedie was generally
providing the best fit to the data. Treatment is used as a parametric term.
According to Gavin, when we use the by-variance smooths, these smooths are
centered around zero (or something effectively similar to that), which means
that we lose information about any mean differences between the treatments.

Before I understood this, I was putting the plots in as random effects, but
when you looked at the diagnostic plots, it looked like the random effects
were basically just treatment effects - which now makes sense. I think the
random effect was basically picking up the mean differences in the treatments.

When we plot these models way down below, I think the plots are telling us that
this model overfits (is overly wiggly), so once we get everything working,
I think it will be worth seeing if we can simplify this model by eliminating
the season-trend interaction. If I understand Gavin's post correctly, however,
I *think* this will make the later steps even more complicated because it
will be harder to find everything we need in the Lp matrix.

```{r}
dipoN_tw = gam(n ~ treatment + te(Time,month, bs = c("cr","cc"), k =c(10,12), by=treatment),
               data = dipoN, family=tw, method = "REML", knots = knots)
summary(dipoN_tw)
```

If I'm interpreting the model summary correctly, on average, the kagnaroo rat 
exclosures (EC) have lower dipo numbers than the rodent exclosures, which are 
lower than the controls on average. The different trend-season fits seem to be really wiggly for the
control plots (CC) and the rodent exclosures (XC).

The plots below emphasize to me the overly qiggly nature of the control 
and rodent exclosure fits.
```{r}
plot(dipoN_tw, pers=TRUE)
```
If we check the gam fit results, that look ok, but not great. It seems worth trying a
different data model for the fit to see if we get any improvement.
```{r}
gam.check(dipoN_tw)
```

Next! See if you can implement the difference part of Gavin's tutorial

## Making the data for the prediction matrix

To both compare the treatments and plot the overall models, we need to 
generate the predicted data. This section just generates the data needed
for the model

```{r}
min_time = min(dipoN$Time)
max_time = max(dipoN$Time)
Time = seq(min_time, max_time, length=1000)
pdat = expand.grid(Time = Time, treatment = as.factor(c("CC","XC","EC")))
pdat_temp = as.Date.numeric(pdat$Time *1000, origin=min('1970-01-01'))
pdat$month = as.numeric(format.Date(pdat_temp, "%m"))
```

The following is Gavin's function to generate the differences between
the treatments. It is lifted from his [blog post](https://www.fromthebottomoftheheap.net/2017/10/10/difference-splines-i/).
It uses the predicted data dataframe to generate a linear predictions matrix (Xp).
The Xp matrix is different from the fittedvalues matrix we use to plot the GAM
model. It contains info about the evaluation of the various basis functions for
the prediction data. Gavins function extracts this info for for each of the 
treatments so they can be compared to each other. 

```{r}
smooth_diff <- function(model, newdata, f1, f2, var, alpha = 0.05,
                        unconditional = FALSE) {
    ## make predicted data
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    ## find columns and rows for treatment pair to be compared
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^te\\(', colnames(xp))] <- 0
    ## Make difference matrix
    dif <- X %*% coef(model)
    # Compute uncertainty in the estimated difference
    se <- rowSums((X %*% vcov(model, unconditional = unconditional)) * X)
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

```

Now we use that function in conjunction with the prediction data to generate
underlying fit information and then plot the difference between the model
fits for the different treatments
```{r}
CC_EC <- smooth_diff(dipoN_tw, pdat, 'CC', 'EC', 'treatment')
CC_XC <- smooth_diff(dipoN_tw, pdat, 'CC', 'XC', 'treatment')
XC_EC <- smooth_diff(dipoN_tw, pdat, 'XC', 'EC', 'treatment')
comp <- cbind(date = Time, rbind(CC_EC, CC_XC, XC_EC))


ggplot(comp, aes(x = as.Date.numeric(date *1000, origin=min('1970-01-01')), y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 2) +
    coord_cartesian(ylim = c(-10,10)) +
    geom_vline(xintercept=as.Date("2014-03-11")) +
    geom_hline(yintercept=0) +
    labs(x = NULL, y = 'Difference in dipoN')

```
This is where is seems like something has gone wrong or I really don't 
understand what is going on. If the y-axis is really difference in dipoN
then I understand why it is positive before the flip (higher dipos on
controls is what we expect), but after the flip, the number is still higher on 
the controls (see 1st graph at beginning of notebook). This seems to be 
indicating that EC have higher dipo numbers after the flip? Why does the CC-XC
comparison has a similar intrpretation issue and why is the standard error
so tight? I have no idea. Ideas: something is wrong with how I pulled out the
treatments from the Xp matrix. Could XC-EC be CC-EC and CC-EC be CC-XC? Do I
misunderstand what the 0 on the y-axis indicate? Is te() smooth causing
issues with Gavin's approach?

SO, this suggests we sould look at the model fits to the data. Here Iattempt to
do this:

```{r}
p  <- predict(dipoN_tw,  newdata = pdat, type="response",se.fit = TRUE)
plotdata <- transform(pdat, fitted=p$fit, se=p$se.fit)
plotdata = dplyr::mutate(plotdata, lower = fitted - 1.96*se, upper = fitted + 1.96*se)

p = ggplot(aes(x=as.Date.numeric(Time *1000, origin=min('1970-01-01')), y=fitted, colour=treatment), data=plotdata) +
    geom_line() +
    geom_point(aes(x=censusdate, y=n, colour=treatment), data=dipoN)
p + geom_ribbon(aes(ymin = lower, ymax = upper), fill='gray90',alpha=.6)

```

To me the overlap jagged nature of the fits (both the predicted values and the
error) suggests overfitting for the controls and rodent exclosures. The 
treatments seem to be in relatively the correct places, so if there is an error
with the treatments for the difference plots then it is coming out in how
I'm pulling the treatments out of the Xp matrix.