---
title: "Dipo_GAM_analysis"
author: EMC 1/2018
output: html_notebook
---



```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)


```

# Portal Plot Flip using gam()

This project is exploring the effect of a resident (animal) community on the 
ability of an additional species to colonize a patch. Things like priority effects 
imply that it should be difficult for a new species to move in to an occupied patch, 
but coexistence theories generally predict that the ability of a community of species 
to coexist and the dominance heirarchy of that community should be independent of who 
got there first.

Main question: does the presence of an established community on a patch slow or prevent 
the "invasion" of a species that should be dominant in the area?

At the Portal Project we have the perfect experimental setup to test this. Kangaroo rats
(genus Dipodomys, of which we have 3 species at Portal) are the clear dominant members of
the rodent community. The Portal Project consists of 24 fenced plots divided into 3 
treatment types: controls (large gates in fences, all rodents have access), kangaroo 
rat exclosures (small gates in fences, kangaroo rats are excluded but all other species 
have access), and total removals (no gates in fences, no rodents have access). 

In March 2015 we changed the treatment on a subset of the plots, cutting new gates or 
enlarging existing gates on restricted plots to allow access to all rodents (including 
kangaroo rats). For this project we're comparing plots that used to be k-rat exclosures 
and became controls (I label these treatment 'EC'), plots that used to be total rodent 
removals and became controls ('XC'), and plots that were always controls and remained 
controls ('CC'). So k-rats are invading an established community in the 'EC' treatment, 
they (along with the other species) are invading an empty patch in the 'XC' treatment, 
and the "background" control community is represented in the 'CC' treatment. 

We're using GAMs to ask whether the density of kangaroo rats on these three treatments 
remains different over time or whether they all converve to the 'CC' eventually. We 
combined all 3 species of dipodomys in this calculation, because all 3 species were 
excluded from the k-rat exclosure plots.

There are a few replicates of each of the three treatment types (3-4 plots per treatment), 
but we decided to take the average of the plots by treatment for each time step rather 
than using plot as a random factor (some of our early model attempts indicated that 
including plot as a factor did not improve models). Also, converting to k-rats/area instead 
of using raw integer counts made the modeling easier (we felt we could use a Gaussian 
family in the GAM). 

## Load Data
Data contains columns for censusdate, numericdate (so we don't have to worry about date 
formatting), treatment, and dipos (units are k-rats per .25 hectare-- we'll probably convert 
this to hectares for the final paper). 


```{r}
# load data. I saved the data series to csv because it takes a few minutes to create them from scratch from create_data_series.R
dipoN_avg = read.csv('Dipo_abundance_by_treatment.csv')

dipoN_avg$censusdate = as.Date(dipoN_avg$censusdate)
dipoN_avg

```


## Basic plot of data

Here's a graph of the data by treatment type. We see the long-term control plots 
(CC, green) are relatively stable through time (dip in most recent months) and the 
increase of the kangaroo rats on the two opened plots after the treatment change, 
with the number of kangaroo rats appearing higher on the opened rodent exclosures 
(XC, purple) than the kangaroo rat exclosures (EC, red). Vertical line is when we 
did the treatment change. (before the treatment change, both EC and XC treatments 
should have zero k-rats. A few show up in the data because sometimes they make it 
into the fences anyway, but they are promptly removed.)


```{r}
ggplot(dipoN_avg, aes(x = censusdate, y =dipos, colour = treatment)) +
    geom_point() +
    geom_smooth(method = 'loess', se = TRUE) +
    scale_colour_brewer(type = 'qual', palette = 'Dark2') +
    theme(legend.position = 'top') +
  geom_vline(xintercept=as.Date('2015-04-10'))
```

## GAM model - ordered factor for treatment type

This is strictly following Gavin's post-- the one on ordered factors
https://www.fromthebottomoftheheap.net/2017/12/14/difference-splines-ii/

```{r pressure, echo=FALSE}
# create ordered factor
dipoN_avg = mutate(dipoN_avg, oTreatment = ordered(treatment, levels = c('CC','EC','XC')))


m <- gam(dipos ~ treatment + s(numericdate) + s(numericdate, by = oTreatment), data = dipoN_avg, method='REML')

plot(m, shade = TRUE, pages = 1,  scale=0,seWithMean = T)

```

This looks weird to me. It looks like the EC treatment is increasing steadily over time, 
but that is not what it looks like from the plot of the data above, nor what we expected 
to happen.

 
```{r pressure, echo=FALSE}
# is this a good model?
summary(m)
gam.check(m)

```

```{r pressure, echo=FALSE}
# Make nice plots
pred = dipoN_avg[,c('numericdate','treatment','oTreatment','censusdate')]
pred$output = predict(m,pred)
pred$se <- predict( m , se = TRUE)$se.fit

pred$lcl <- pred$output - 1.96 * pred$se
pred$ucl <- pred$output + 1.96 * pred$se
ggplot(pred) +
  geom_line(aes(x=censusdate,y=output,colour=treatment)) +
  geom_ribbon(aes(x=censusdate,ymax=ucl,ymin=lcl,fill=treatment),alpha=.5) +
  geom_vline(xintercept=as.Date('2015-04-10'))
  

```
This plot seems useful: we can talk about the difference in treatments by comparing overlap 
of the confidence intervals. However this plot doesn't look anything like the "difference" 
plots in the section above -- why??