---
title: "Rodent Flip Analysis PT2"
output: html_notebook
---

# Portal Rodent Flip: Fitting Gams using Gaussian Fit as Proof of Concept

This notebook is working through the approach to analyzing the rodent flip
data using GAMS given what Erica and I have learned from Simon Wood's book and
Gavin Simpson's blog posts. Even though Gaussian is not the correct family
to use, it is the easiest and the one all the examples are in, so this is
a proof of concept on how to approach this question generally and then we will
explore if its possible to do this with other distributions.

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)
source('gam_functions.R')
source('data_functions.R')
```
## Load and Vizualize Data

Data is coming straight from the repo. This code is extracting the dipo a
bundances from the straight controls (CC), and the opened krat (EC) and rodent 
(XC) exclosures. This data runs from period 415 (3/13/2013) to present. We
have not summarized the data by treatment - each plot is represented individually.

```{r}
dipo_data = make_dipo_data()
filtered_data = trt_data(dipo_data)
CC = filtered_data[[1]]
EC = filtered_data[[2]]
XC = filtered_data[[3]]
```
## Control Plots

To assess basic data family, I'm going to use a basic model structure:

$$ DipoN = s(Time) + s(season), DipoN ~ Gaussian$$

This structure has no interaction, no AR. A cc basis is used to account for
seasonality, but this also **does not fix the knot problem Gavin highlighted**. 
I'm starting in gam() because the two other distributions don't work in gamm()
unless their parameters have already been estimated in gam().

Model names will take structure: 
treatment_family_AR
if an interaction is being modelled, and X will be added at the end.

### Basic Model with gam()

```{r}

knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_0 <- gam(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, knots=knots)
gam.check(cc_0)
```

The qqplot is off and there might be some patterning in the residuals. So,
poisson is probably not our best distribution, but we're going to run
with it because the other two contenders are....difficult.

In slides on the web 
(https://statistique.cuso.ch/fileadmin/statistique/document/part-3.pdf), Simon 
Wood recommends some additional plots for checking model fit.

```{r}
rsd <- residuals(cc_0)
qq.gam(cc_0,rep=100); 
plot(cc_0,residuals=TRUE,pch=19,cex=.3)
```
Yep, the qq-plot is definitely off, even when the distribution is simulated.

```{r}
plot(cc_0,residuals=TRUE,pch=19,cex=.3)
```

The residuals around the time smooth also show some systematic deviations. The
season graph looks like some years are extremely high, relative to the others.


For later comparisonwith gamm(), lets look at the summary:

```{r}
summary(cc_0)
```

Season and Trend are both significant in this model. 

The next step is to add autocorrelation, but to do that we need to switch to
gamm()

### Basic control plot model with gamm()

There are differences between gam() and gamm() in fitting and (I think)
penalities, so let's do this again but with gamm()


```{r}
old.par=par()
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_0 <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
               knots=knots, method="REML")
gam.check(cc_0$gam)
```

looks pretty similar.

Model summary:

```{r}
summary(cc_0$gam)
```
This fit is slightly lower than the gam() fit. 

### Adding Autrocorrelation to the model

Ok, let's proceed with Gavin's excercise with adding AR to the model. Let's
look at the autocorrelation pattern.

```{r}
plot(acf(resid(cc_0$lme, type = "normalized")))
```
```{r}
plot(pacf(resid(cc_0$lme, type = "normalized")))
```


It's a strange autocorrelation structure for Portal rodents. In part this may be 
because themodel picks up (absorbs) some of the wiggliness introduced by the 
autocorrelationbetween months. Lag 2 or lag 4 may be important. It's also 
possible that thispattern suggests a MA not an AR model should be applied to the 
residuals?

Let' start with AR, using Gavin's code/approach from:[insert webpage]. These are 
within year autocorrelation structures only. Given the biology of this system, 
I think that's a good starting point. With El Nino bouncing things around, and 
the natural between year variability of the desert, most of the autocorrelation 
is likely to be on the month-to-month scale.


```{r}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)
for (i in 1:8) {
    cc <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
               control = ctrl, knots=knots, method="REML",
               correlation = corARMA(form = ~ 1 | Year, p = i))
    assign(paste0("cc_", i), cc) 
}
```

```{r}

anova(cc_0$lme, cc_1$lme,cc_2$lme,cc_3$lme,cc_4$lme,cc_5$lme,
      cc_6$lme,cc_7$lme, cc_8$lme)

```

Looks like AR2 is better than AR1, AR5 is might be better than AR4. The p-value
for AR5 is slight above .05 and the AIC value is higher. GIven the ACF and PACF
I'm going to ignore AR5. But is AR2 better than AR4?

```{r}
anova(cc_2$lme,cc_4$lme)
```

AR4 it is. Let's look at this model:

```{r}
gam.check(cc_4$gam)
```

```{r}
summary(cc_4$gam)
```


The adj r2 is lower, the trend component is now non-significant. i.e. lots
of variation in the dipo numbers on the control plots but little of it
explained by time. This is not surprising. 

```{r}
summary(cc_4$lme)
```

What do the ACF/PACF plots look like now?

```{r}
plot(acf(resid(cc_4$lme, type = "normalized")))
```

```{r}
plot(pacf(resid(cc_4$lme, type = "normalized")))
```

Everything looks good.

ok, then given a gaussian distribution - model cc_4 is our control fit.

```{r}
plot(cc_4$gam)
```

This model is telling us that on the controls, the number of kangaroo rats is
slightly higher in the winters and pretty flat across years.

### Season-Trend interaction?

Ok, first we need to construct a season-trend interaction model. We'll be doing
the AR fitting again as part of this excercise. it will fit the main and
interaction effects of the seasonal and trend signals in the control data, using
the knot structure for the monthly data that Gavin suggests.

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
for (i in 1:5) {
  cc_X <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),                data = CC, method = "REML", control = ctrl, 
               knots = knots, 
               correlation = corARMA(form = ~ 1 | Year, p = i))
  assign(paste0("cc_X_", i), cc_X) 
}
anova(cc_X_1$lme, cc_X_2$lme, cc_X_3$lme, cc_X_4$lme, cc_X_5$lme)
```

Looks like CC_X_4 generally does best via AIC measures.

Let's look at the ACF/PACF plots.

```{r}
plot(acf(resid(cc_X_4$lme, type = "normalized")))
```

```{r}
plot(pacf(resid(cc_X_4$lme, type = "normalized")))
```

Looks good. The summary:

```{r}
summary(cc_X_4$gam)
```

Seems like everytime I run a model, the adjR2 gets lower and lower. The
te() smooth is not significant. Which this plot seems to support:

```{r}
plot(cc_X_4$gam, pers = TRUE)
```

There doesn't seem to be much of an interaction effect here. To me, this looks like how the number of kangaroo rats on the control plots varies through a year is not changing across years.

Do we need the interaction (which seems to make the fit worse overall anyway?). I think the answer from this excercise alone is hell no.

### Plots as Random Effect

Since the plots are just samples representing the variance of responses and not
of interest themselves, I'm putting the plots in as random effects.

```{r}
CC$plot = as.factor(CC$plot)
cc_4_re <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
                control = ctrl, knots=knots, method="REML", random=list(plot=~1),
                correlation = corARMA(form = ~ 1 | Year, p = 4))
anova(cc_4_re$lme,cc_4$lme)
```
Looks like we're ok with not incorporating the plots - there's no support for
accepting a model that models mean differences in the residuals among the plots.

### Model Checking

Basic residual checking was done above. Here are some additional plots:

```{r}
plot(cc_4$gam,residuals=T,pch=19)
```
There's still some patterning with respect to time that the model is not getting.

There may be variance changes with month and definitely some wiggles in the
data that aren't coming out with the AR or time smooth. So this is not a great
model - but how much is because the Gaussian just isn't a great model for this
discrete data? I don't know.

We can see how the model looks on the data. Is this something I would want
to predict the future from? No, but it is not an unreasonable facsimile of the
control temporal pattern.

```{r}
  # Make data to plot the trend line on the data
times = select(CC,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(cc_4$gam,  newdata = pdat, se.fit = TRUE)
pdat <- transform(pdat, fitted=p$fit, se=p$se.fit)
pdat = pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p1 = ggplot(aes(x = date, y = fitted), data = pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'blue') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("CC_4 model") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
p1 + geom_point(data=CC, aes(x=date,y=DipoN))
```

## Kangaroo Rat Exclosures






11