---
title: "Rodent Flip Analysis PT2"
output: html_notebook
---

# Portal Rodent Flip: Fitting Gams using Gaussian Fit as Proof of Concept

This notebook is working through the approach to analyzing the rodent flip
data using GAMS given what Erica and I have learned from Simon Wood's book and
Gavin Simpson's blog posts. Even though Gaussian is not the correct family
to use, it is the easiest and the one all the examples are in, so this is
a proof of concept on how to approach this question generally and then we will
explore if its possible to do this with other distributions.

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)
source('gam_functions.R')
source('data_functions.R')
```
## Load Data

Data is coming straight from the repo. This code is extracting the dipo 
abundances from the straight controls (CC), and the opened krat (EC) and rodent 
(XC) exclosures. This data runs from period 415 (3/13/2013) to present. We
have not summarized the data by treatment - each plot is represented individually.

```{r}
dipo_data = make_dipo_data()
filtered_data = trt_data(dipo_data)
CC = filtered_data[[1]]
EC = filtered_data[[2]]
XC = filtered_data[[3]]
```
## Control Plots

```{r}
plot(DipoN ~ date, data=CC, col='blue')
```


As a starting point, we're going to use a basic time model structure:

$$ DipoN = s(Time) + s(season), DipoN  Gaussian$$

This structure has no interaction, no AR. A cc basis is used to account for
seasonality. We're starting with a Gaussian - even though it is unlikely to be
the distribution family we want to use in the final product but it is an easy
distribution to start implementing. Everything else is more...complicated. But
our data is discrete count data, so Gaussian is not really appropriate
.
I'm starting in gam() because the two other distributions don't work in gamm()
unless their parameters have already been estimated in gam().

Model names will take structure: 
treatment_family_AR
if an interaction is being modelled, and X will be added at the end.

### Basic Model with gam()

```{r}

knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_0 <- gam(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, knots=knots)
gam.check(cc_0, rep=100)
```

The qqplot is off and there might be some patterning in the residuals. So,
poisson is probably not our best distribution, but we're going to run
with it because the other two contenders are....difficult.

In slides on the web 
(https://statistique.cuso.ch/fileadmin/statistique/document/part-3.pdf), Simon 
Wood recommends some additional plots for checking model fit.

```{r}
rsd <- residuals(cc_0)
qq.gam(cc_0,rep=100); 
plot(cc_0,residuals=TRUE,pch=19,cex=.3)
```
Yep, the qq-plot is definitely off, even when the distribution is simulated.

```{r}
plot(cc_0,residuals=TRUE,pch=19,cex=.3)
```

The residuals around the time smooth also show some systematic deviations. The
season graph looks like some years are extremely high, relative to the others.


For later comparisonwith gamm(), lets look at the summary:

```{r}
summary(cc_0)
```

Season and Trend are both significant in this model. 

The next step is to add autocorrelation, but to do that we need to switch to
gamm()

### Basic control plot model with gamm()

There are differences between gam() and gamm() in fitting and (I think)
penalities, so let's do this again but with gamm()


```{r}
old.par=par()
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_0 <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
               knots=knots, method="REML")
gam.check(cc_0$gam)
```

looks pretty similar.

Model summary:

```{r}
summary(cc_0$gam)
```
This fit is slightly lower than the gam() fit. 

### Adding Autrocorrelation to the model

Ok, let's proceed with Gavin's excercise with adding AR to the model. Let's
look at the autocorrelation pattern.

```{r}
plot(acf(resid(cc_0$lme, type = "normalized")))
```
```{r}
plot(pacf(resid(cc_0$lme, type = "normalized")))
```


It's a strange autocorrelation structure for Portal rodents. In part this may be 
because themodel picks up (absorbs) some of the wiggliness introduced by the 
autocorrelationbetween months. Lag 2 or lag 4 may be important. It's also 
possible that thispattern suggests a MA not an AR model should be applied to the 
residuals?

Let' start with AR, using Gavin's code/approach from:[insert webpage]. These are 
within year autocorrelation structures only. Given the biology of this system, 
I think that's a good starting point. With El Nino bouncing things around, and 
the natural between year variability of the desert, most of the autocorrelation 
is likely to be on the month-to-month scale.


```{r}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)
for (i in 1:8) {
    cc <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
               control = ctrl, knots=knots, method="REML",
               correlation = corARMA(form = ~ 1 | Year, p = i))
    assign(paste0("cc_", i), cc) 
}
```

```{r}

anova(cc_0$lme, cc_1$lme,cc_2$lme,cc_3$lme,cc_4$lme,cc_5$lme,
      cc_6$lme,cc_7$lme, cc_8$lme)

```

Looks like AR2 is better than AR1, AR5 is might be better than AR4. The p-value
for AR5 is slight above .05 and the AIC value is higher. GIven the ACF and PACF
I'm going to ignore AR5. But is AR2 better than AR4?

```{r}
anova(cc_2$lme,cc_4$lme)
```

AR4 it is. Let's look at this model:

```{r}
gam.check(cc_4$gam)
```

```{r}
summary(cc_4$gam)
```


The adj r2 is lower, the trend component is now non-significant. i.e. lots
of variation in the dipo numbers on the control plots but little of it
explained by time. This is not surprising. 

```{r}
summary(cc_4$lme)
```

What do the ACF/PACF plots look like now?

```{r}
plot(acf(resid(cc_4$lme, type = "normalized")))
```

```{r}
plot(pacf(resid(cc_4$lme, type = "normalized")))
```

Everything looks good.

ok, then given a gaussian distribution - model cc_4 is our control fit.

```{r}
plot(cc_4$gam)
```

This model is telling us that on the controls, the number of kangaroo rats is
slightly higher in the winters and pretty flat across years.

### Season-Trend interaction?

Ok, first we need to construct a season-trend interaction model. We'll be doing
the AR fitting again as part of this excercise. it will fit the main and
interaction effects of the seasonal and trend signals in the control data, using
the knot structure for the monthly data that Gavin suggests.

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
for (i in 1:5) {
  cc_X <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),                data = CC, method = "REML", control = ctrl, 
               knots = knots, 
               correlation = corARMA(form = ~ 1 | Year, p = i))
  assign(paste0("cc_X_", i), cc_X) 
}
anova(cc_X_1$lme, cc_X_2$lme, cc_X_3$lme, cc_X_4$lme, cc_X_5$lme)
```

Looks like CC_X_4 generally does best via AIC measures.

Let's look at the ACF/PACF plots.

```{r}
plot(acf(resid(cc_X_4$lme, type = "normalized")))
```

```{r}
plot(pacf(resid(cc_X_4$lme, type = "normalized")))
```

Looks good. The summary:

```{r}
summary(cc_X_4$gam)
```

Seems like everytime I run a model, the adjR2 gets lower and lower. The
te() smooth is not significant. Which this plot seems to support:

```{r}
plot(cc_X_4$gam, pers = TRUE)
```

There doesn't seem to be much of an interaction effect here. To me, this looks like how the number of kangaroo rats on the control plots varies through a year is not changing across years.

Do we need the interaction (which seems to make the fit worse overall anyway?). I think the answer from this excercise alone is hell no.

### Plots as Random Effect

Since the plots are just samples representing the variance of responses and not
of interest themselves, I'm putting the plots in as random effects.

```{r}
CC$plot = as.factor(CC$plot)
cc_4_re <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, 
                control = ctrl, knots=knots, method="REML", random=list(plot=~1),
                correlation = corARMA(form = ~ 1 | Year, p = 4))
anova(cc_4_re$lme,cc_4$lme)
```
Looks like we're ok with not incorporating the plots - there's no support for
accepting a model that models mean differences in the residuals among the plots.

### Model Checking

Following Simon Wood notes from: https://statistique.cuso.ch/fileadmin/statistique/document/part-3.pdf

```{r}
vis.gam(cc_4$gam)
```
There's still some patterning with respect to time that the model is not getting.

There may be variance changes with month and definitely some wiggles in the
data that aren't coming out with the AR or time smooth. So this is not a great
model - but how much is because the Gaussian just isn't a great model for this
discrete data? I don't know.

We can see how the model looks on the data. Is this something I would want
to predict the future from? No, but it is not an unreasonable facsimile of the
control temporal pattern.

```{r}
  # Make data to plot the trend line on the data
times = select(CC,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
cc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(cc_4$gam,  newdata = cc_pdat, se.fit = TRUE)
cc_pdat <- transform(cc_pdat, fitted=p$fit, se=p$se.fit)
cc_pdat = cc_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p1 = ggplot(aes(x = date, y = fitted), data = cc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'blue') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("CC_4 model") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
p1 + geom_point(data=CC, aes(x=date,y=DipoN))
```

## Kangaroo Rat Exclosures

Ok, here's what the data for the Kangaroo Rat Exclosures looks like:

```{r}
plot(DipoN ~ date, data=EC, col='green')
```
### Basic Model

Here are a bunch of diagnostic plots for the basic model that includes the
season and time effects as additive and does not include an AR model.

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
ec_0 <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = EC, knots=knots,
             method="REML")
gam.check(ec_0$gam)
```

```{r}
rsd <- residuals(ec_0)
qq.gam(ec_0$gam,rep=100); 
plot(ec_0$gam,residuals=TRUE,pch=19,cex=.3)
```

The residuals deviate clearly from Gaussian. We have more zeros than we did 
before. Clearly a Gaussian would not be anyone's first choice for this data.

### Autoregressive Models

```{r}
plot(acf(resid(ec_0$lme, type = "normalized")))
```
```{r}
plot(pacf(resid(ec_0$lme, type = "normalized")))
```

```{r}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)
for (i in 1:8) {
    ec <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = EC, 
               control = ctrl, knots=knots, method="REML",
               correlation = corARMA(form = ~ 1 | Year, p = i))
    assign(paste0("ec_", i), ec) 
}
anova(ec_0$lme, ec_1$lme, ec_2$lme, ec_3$lme, ec_4$lme, ec_5$lme)
```
```{r}
anova(ec_0$lme, ec_3$lme,ec_5$lme)
```
I think this is tell me that theres not much difference between these models. 

I suspect that is because of the first half of the time series, the dependence
of one month on the previous ones is supressed with the exclosures, and becomes
re-established when they are opened to controls. We'll see what the interactive
model says.

### Season-Trend Interaction?

AR3 has convergence issues, for reasons I don't understand. Could this be due to the fact that Gaussian is clearly wrong for this data? Or the high number of zeros? But then why to the others run? Here are the models that do run:

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))

ec_X_0 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),
               data = EC, method = "REML", control = ctrl, knots = knots)

ec_X_1 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)), 
               data = EC, method = "REML", control = ctrl, knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 1))
ec_X_2 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),
               data = EC, method = "REML", control = ctrl, knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 2))
ec_X_4 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),
               data = EC, method = "REML", control = ctrl, knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 4))
ec_X_5 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),
               data = EC, method = "REML", control = ctrl, knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 5))

anova(ec_X_0$lme, ec_X_1$lme, ec_X_2$lme, ec_X_4$lme, ec_X_5$lme)
```
```{r}
anova(ec_X_0$lme, ec_X_5$lme)
```

Looks like the AR5 interaction model wins. Let's look at it to see if its worth
assessing whether it is better than a strictly additive model.

```{r}
summary(ec_X_5$gam)
```

Huh, the interaction is singificant and the r2 isn't abyssmal.

Let's look at the residual plots:

```{r}
gam.check(ec_X_5$gam)
```

Well, none of those look awesome. Clearly variance increases with mean, the
residual histogram and qqplot is....odd. Gaussian still sucks.

Let's see if the interaction is better than the additive model:

```{r}
phi <- unname(intervals(ec_X_5$lme, which = "var-cov")$corStruct[, 2])
ec_X_5_test <- gamm(DipoN ~ s(Time, bs = "cr", k = 10) 
                    + s(month, bs = "cc",k = 12) 
                    + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
                    data = EC, method = "ML", control = ctrl, knots = knots,
                    correlation = corARMA(value = phi, fixed = TRUE, form = ~ 1 | 
                                            Year, p = 5))
ec_5_test <- gamm(DipoN ~ s(Time, bs = "cr", k = 10) 
                  + s(month, bs = "cc",k = 12), 
                  data = EC, method = "ML", control = ctrl, knots = knots,
                  correlation = corARMA(value = phi, fixed = TRUE, 
                                        form = ~ 1 | Year, p = 5))
anova(ec_X_5_test$lme, ec_5_test$lme)
```

Looks like the interaction is a pretty substantive improvement.


### Adding plots as random effect
```{r}
EC$plot = as.factor(EC$plot)
ec_X_5_plot <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),
               data = EC, method = "REML", control = ctrl, knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 5), 
               random=list(plot= ~1 ))
anova(ec_X_5$lme, ec_X_5_plot$lme)
```
Looks like the simpler model without plots as random effects is fine.

### Model Checking

```{r}
plot(ec_X_5$gam,residuals=T,pch=19, pers = TRUE)
```
```{r}
times = select(EC,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
ec_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(ec_X_5$gam,  newdata = ec_pdat, se.fit = TRUE)
ec_pdat <- transform(ec_pdat, fitted=p$fit, se=p$se.fit)
ec_pdat = ec_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p1 = ggplot(aes(x = date, y = fitted), data = ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'blue') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("EC_X_5 Season-trend Interaction model") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
p1 + geom_point(data=EC, aes(x=date,y=DipoN))
```

## Rodent Exclosures

```{r}
plot(DipoN ~ date, data=XC, col='red')
```

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
xc_0 <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = XC, knots=knots,
             method="REML")
gam.check(xc_0$gam)
```
```{r}
rsd <- residuals(xc_0)
qq.gam(xc_0$gam,rep=100); 
plot(xc_0$gam,residuals=TRUE,pch=19,cex=.3)
```

In summary, the basic tim model with Gaussian family is performing no worse
with this data than it did with the other treatment types.

### Autoregressive Residuals

```{r}
plot(acf(resid(xc_0$lme, type = "normalized")))
```

```{r}
plot(pacf(resid(xc_0$lme, type = "normalized")))
```

Maybe a slight hint of a AR6? But in general, this is pretty good looking - in
large part no doubt because when we excluded rodents, we were completely
disrupting the month to month dependency.

```{r}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)
for (i in 1:7) {
    xc <- gamm(DipoN ~ s(month, bs = "cc", k = 12) + s(Time), data = XC, 
               control = ctrl, knots=knots, method="REML",
               correlation = corARMA(form = ~ 1 | Year, p = i))
    assign(paste0("xc_", i), xc) 
}
anova(xc_0$lme, xc_1$lme, xc_2$lme, xc_3$lme, xc_4$lme, xc_5$lme, xc_6$lme, xc_7$lme )
```

```{r}
anova(xc_0$lme, xc_3$lme, xc_6$lme)
```

AR6? 

### Season-Trend Interaction?

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
xc_X_0 <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = XC, method = "REML", control = ctrl, knots = knots)
for (i in 1:6) {
  xc_X <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = XC, method = "REML", control = ctrl, knots = knots, 
               correlation = corARMA(form = ~ 1 | Year, p = i))
  assign(paste0("xc_X_", i), xc_X) 
}
anova(xc_X_0$lme, xc_X_1$lme, xc_X_2$lme, xc_X_3$lme, xc_X_4$lme, xc_X_5$lme, xc_X_6$lme)
```

```{r}
anova(xc_X_1$lme, xc_X_4$lme)
```


AR4 model when the interaction is included. 

```{r}
# phi <- unname(intervals(xc_X_4$lme, which = "var-cov")$corStruct[, 2])
# xc_X_4_test <- gamm(DipoN ~ s(Time, bs = "cr", k = 10) 
#                     + s(month, bs = "cc",k = 12) 
#                     + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
#                     data = XC, method = "ML", control = ctrl, knots = knots,
#                     correlation = corARMA(value = phi, fixed = TRUE, form = ~ 1 | 
#                                             Year, p = 5))
# xc_4_test <- gamm(DipoN ~ s(Time, bs = "cr", k = 10) 
#                   + s(month, bs = "cc",k = 12), 
#                   data = XC, method = "ML", control = ctrl, knots = knots,
#                   correlation = corARMA(value = phi, fixed = TRUE, 
#                                         form = ~ 1 | Year, p = 5))
# anova(ec_X_5_test$lme, ec_5_test$lme)
```
As with the control plot, it is not possible to extract the AR 
cofficients using Gavin's code. When I googled the error it suggests
the model is inappropriate for the data - which could be the Gaussian
issue with discrete data? This makes it impossible for me to test
whether the interaction model is better than the additive model.

### Plots as random effect


To complete the exercise, I'm going to go ahead and look at the interaction model. 
```{r}
XC$plot = as.factor(XC$plot)
xc_X_4_plot <- gamm(DipoN ~ te(Time,month, bs = c("cr","cc"), 
                               k=c(10,12)),
               data = XC, method = "REML", control = ctrl, 
               knots = knots,
               correlation = corARMA(form = ~ 1 | Year, p = 4), 
               random=list(plot= ~1 ))
anova(xc_X_4$lme, xc_X_4_plot$lme)
```
No plot effect here either.

#### Model checking

We'll use xc_X_4 for this, but with the understanding that given
the issues with fitting, it should be viewed even more suspiciously
than the other models.

```{r}
summary(xc_X_4$gam)
```

The ref edf and the edf are exactly the same. I should check Wood about
whether I need to increase k in the model.

```{r}
plot(xc_X_4$gam,residuals=T,pch=19, pers = TRUE)
```


The rodent exclosure seems to have a more complex surface.

```{r}
times = select(XC,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
xc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(xc_X_4$gam,  newdata = xc_pdat, se.fit = TRUE)
xc_pdat <- transform(pdat, fitted=p$fit, se=p$se.fit)
xc_pdat = pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_xc = ggplot(aes(x = date, y = fitted), data = xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'red') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("EC_X_5 Season-trend Interaction model") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
p_xc + geom_point(data=XC, aes(x=date,y=DipoN))
```

So, what happens if I plot the three lines together?

```{r}
transition = as.Date("2015-03-15", format="%Y-%m-%d")
plot_all = ggplot(aes(x = date, y = fitted), data = xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'red') +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = ec_pdat, fill = 'gray90') +
  geom_line(color = 'blue', data=ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = cc_pdat, fill = 'gray90') +
  geom_line(color = 'green', data=cc_pdat) +
  geom_vline(xintercept =  as.numeric(as.Date('2015-03-20'))) +
  ggtitle("Dipodomys response to plot flip") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
  
plot_all



```

