---
title: "GAM_nongaussian"
output: html_notebook
---

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)

source('gam_functions.R')
source('data_functions.R')
```

# Portal Plot Flip using gam()

We have to use gamm() if we are modelling the AR in the data, but it is also a more advanced tool that is easier to use incorrectly. We only currently knopw how to use it with a Gaussian family distribution and even the experts on the stats sites seem to advice trying something other than non-gaussian with AR in gamm().

The Gaussian family is a terrible fit to the data - which you would expect from count data. So here we're ditching the AR structure in favor of using a more justifiable distribution.

Here we go:

## Load Data

Data is coming straight from the repo. This code is extracting the dipo abundances from the straight controls (CC), and the opened krat (EC) and rodent (XC) exclosures. This data runs from period 415 (3/13/2013) to present

```{r}
mass_data = get_mass(start_period = 415,avg=T,metE=F) %>% trt_data()

CC = mass_data[[1]]
EC = mass_data[[2]]
XC = mass_data[[3]]
```

## Checking different distributions
If you don't want to go through all the distribution checking, tweedie seems
to work reasonably for all the treatments. Poisson does not work for the
exclosures. Negative binomial is probably fine but things look slight better
fit with tweedies. So, if you want, you can skip down to the section on
**checking plot effects**

### Control Plots
#### Gaussian
We'll start where we left off for a baseline of comparison: Gaussian 

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_g <- gam(n ~ s(month, bs = "cc", k = 12) + s(Time), data = CC, knots=knots)
gam.check(cc_g, rep=100)
```

```{r}
summary(cc_g)
```
```{r}
plot(cc_g, residuals = TRUE)
```
Gaussian looks like a reasonable starting place. 

`




Seems like poisson and tweedie are fairly similar in fits. Poisson misses more
on the high values, seems to miss more on the low values.

### Kangaroo rat exclosures
#### Gaussian
```{r}

```

#### Poisson
To speed this up, I'm jumping straight to the Poisson.
```{r}
ec_g = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), data = EC, knots=knots)
gam.check(ec_g, rep=100)
```

```{r}
ec_p = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), family=poisson, data = EC, knots=knots)
gam.check(ec_p, rep=100)
```

Poisson doesn't look so good either. 

```{r}
ec_nb = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), family=nb, data = EC, knots=knots)
gam.check(ec_nb, rep=100)
```

Negative Binomial - still not good.

#### Tweedie

```{r}
ec_tw = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), family=tw, data = EC, knots=knots)
gam.check(ec_tw, rep=100)
```
```{r}
plot(ec_tw, residuals = TRUE, cex=5)
```
```{r}
summary(ec_tw)
```

Tweedie seems to win most consistent with the data, but with caveats.

### Rodent Exclosure

Last but not least

#### Gaussian
```{r}
xc_g = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), data = XC, knots=knots)
gam.check(xc_g, rep=100)
```
seems - ok
#### Poisson

```{r}
xc_p = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), data = XC, knots=knots)
gam.check(xc_p, rep=100)
```

#### Negative Binomial

```{r}
xc_nb = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), family=nb, data = XC, knots=knots)
gam.check(xc_nb, rep=100)

```

#### Tweedie

```{r}
xc_tw = gam(n ~ s(month, bs = "cc", k = 12) + s(Time), family=tw, data = XC, knots=knots)
gam.check(xc_tw, rep=100)
```
Gaussian wins.

## Season-Trend interactions
Control plots:

```{r}
cc_g_X = gam(n ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = CC, method = "REML", knots = knots)
summary(cc_g_X)
```
```{r}
plot(cc_g_X, pers=TRUE)
```
Kangaroo rat exclosures:
```{r}
ec_tw_X = gam(n ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = EC, family=tw, method = "REML", knots = knots)
summary(ec_tw_X)
```
```{r}
plot(ec_tw_X, pers=TRUE)
```
Rodent exclosures:

```{r}
xc_g_X = gam(n ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = XC, method = "REML", knots = knots)
summary(xc_g_X)
```
```{r}
plot(xc_tw_X, pers=TRUE)
```
Testing each of the models to see if its better than additive

```{r}
cc_g_X_test <- gam(n ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12) 
                     + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
                     data = CC, method = "REML", knots = knots)
cc_g_test <- gam(n ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12), data = CC, 
                  method = "REML", knots = knots)
AIC(cc_g_X_test, cc_g_test)
anova(cc_g_X_test, cc_g_test)
```
Look like there is nothing gained with the interaction for control plots

Kangaroo rat exclosures?

```{r}
ec_tw_X_test <- gam(n ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12) 
                     + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
                     data = EC, method = "ML", knots = knots, family=tw)
ec_tw_test <- gam(n ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12), data = EC, 
                  method = "ML", knots = knots, family = tw)
AIC(ec_tw_X_test, ec_tw_test)
anova(ec_tw_test,ec_tw_X_test)
```
```{r}
xc_tw_X_test <- gam(n ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12)
                     + ti(Time, month, bs = c("cr","cc") , k = c(10, 12)), 
                     data = XC, method = "ML", knots = knots)
xc_tw_test <- gam(n ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12) , data = XC, 
                  method = "ML", knots = knots)
AIC(xc_tw_X_test, xc_tw_test)
anova(xc_tw_test,xc_tw_X_test)
```
No sign that there is a season-year interaction for average size for any of the treatments.

Interesting.

## Checking Plot effects

Whatever the best model was coming out of the  The
random effect for plot is added as s(plot, bs="re") and that term is checked for
significance.

# Control Plots
```{r}
CC$plot = as.factor(CC$plot)
cc_g_plot = gam(n ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot, bs="re"), data = CC, knots=knots)
summary(cc_g_plot)
```

Plot effect in average size.

```{r}
EC$plot = as.factor(EC$plot)
ec_tw_plot = gam(n ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot, bs="re"),            
               data = EC, family=tw, method = "REML", knots = knots)
summary(ec_tw_plot)
```


```{r}
XC$plot = as.factor(XC$plot)
xc_g_plot = gam(n ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot,bs="re"), data = XC, knots=knots)
summary(xc_g_plot)
```
Plots as a random effect are significant in all of the models.

## PLOTS!
Control with data:
```{r}
times = select(CC,date,month,Time, plot) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
cc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want], plot=plot[want]))
p  <- predict(cc_g_plot,  newdata = cc_pdat, type="response",se.fit = TRUE)
cc_pdat <- transform(cc_pdat, fitted=p$fit, se=p$se.fit)
cc_pdat = cc_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_cc = ggplot(aes(x = date, y = fitted), data = cc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'green') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Controls: Mass ~ s(month) + s(Time) + s(plot, bs=re)") +
  xlab("Date") + ylab("Average mass per plot") +
  theme_classic()
p_cc + geom_point(data=CC, aes(x=date,y=n))
```
Kangaroo rat exclosures with data
```{r}
times = select(EC,date,month,Time, plot) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
ec_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want], plot=plot[want]))
p  <- predict(ec_tw_plot,  newdata = ec_pdat, type="response",se.fit = TRUE)
ec_pdat <- transform(ec_pdat, fitted=p$fit, se=p$se.fit)
ec_pdat = ec_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_ec = ggplot(aes(x = date, y = fitted), data = ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'blue') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Kangaroo rat exclosures: Mass ~ s(month) + s(time)+s(plot,'re'), tweedie") +
  xlab("Date") + ylab("Average mass per plot") +
  theme_classic()
p_ec + geom_point(data=EC, aes(x=date,y=n))
```

```{r}
times = select(XC,date,month,Time, plot) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
xc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want], plot=plot[want]))
p  <- predict(xc_g_plot,  newdata = xc_pdat, type="response",se.fit = TRUE)
xc_pdat <- transform(xc_pdat, fitted=p$fit, se=p$se.fit)
xc_pdat = xc_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_xc = ggplot(aes(x = date, y = fitted), data = xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'red') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Rodent Exclosures: DipoN ~ s(month) + s(Time), tweedie") +
  xlab("Date") + ylab("Dipodomys abundance per plot") +
  theme_classic()
p_xc + geom_point(data=XC, aes(x=date,y=n))
```


```{r}
transition = as.Date("2015-03-15", format="%Y-%m-%d")
ggplot(aes(x=date, y=fitted), data=xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') + 
  geom_line(color = 'red') +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = ec_pdat, fill = 'lightsteelblue2') +
  geom_line(color = 'blue', data=ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = cc_pdat, fill = 'palegreen') +
  geom_line(color = 'green', data=cc_pdat) +
  geom_vline(xintercept =  as.numeric(as.Date('2015-03-20'))) +
  ggtitle("Average Mass Response to FLip") +
  xlab("Date") + ylab("Average Mass") +
  theme_classic()
```


### Comparing two surfaces:

I'm trying to do this following Wood (Section 7.2.5 in his book). 

I'm going to combine the XC and EC datasets into 1 data frame and add some
dummy variables to identify which treatment each comes from.

```{r}
EC_XC=c()
EC_XC = rbind(XC, EC)
EC_XC$XC = c(rep(1,nrow(XC)), rep(0,nrow(XC)))
EC_XC$EC = 1- EC_XC$XC
```

Ok, now here's his approach of fitting a common surface and difference
surface:

```{r}
same = gam(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = EC_XC, family=tw, method = "REML", knots = knots)
diff = gam(DipoN ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)) +
             te(Time,month, by=XC, bs = c("cr","cc"), k =c(10,12)),            
               data = EC_XC, family=tw, method = "REML", knots = knots)
AIC(same,diff)
```
```{r}
anova(diff)
```
```{r}
summary(diff)
```
```{r}
plot(diff, pers=TRUE)
```

So, my read on this is that the difference between the two treatments is high at
the beginning but decreases through time. 

The diff component if significant, but the AIC suggests that it is not a better
fit given the number of parameters. But then these two treatments may not really
differ in a simple way like Wood described for the use of this technique. 
```{r}

```
