---
title: "GAM_nongaussian_sp_rich"
output: html_notebook
---

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)

source('gam_functions.R')
source('data_functions.R')
```

# Portal Plot Flip using gam() -- species richness

Here we are repeating the process we went through with the dipo abundance data to model species richness.  Again, we're using gam() instead of gamm() so we can compare models using different distributions.

Here we go:

## Load Data

Data is coming straight from the repo. This code is extracting the number of species caught from the straight controls (CC), and the opened krat (EC) and rodent (XC) exclosures. This data runs from period 415 (3/13/2013) to present

```{r}
sprich = rodent_abundance(start_period=415,incomplete=F) %>% species_rich() %>% trt_data()
CC_SR = sprich[[1]]
EC_SR = sprich[[2]]
XC_SR = sprich[[3]]
```

## Checking different distributions
If you don't want to go through all the distribution checking, tweedie seems
to work reasonably for all the treatments. Poisson does not work for the
exclosures. Negative binomial is probably fine but things look slight better
fit with tweedies. So, if you want, you can skip down to the section on
**checking plot effects**

### Control Plots
#### Gaussian
We'll start where we left off for a baseline of comparison: Gaussian 

```{r}
knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
cc_g <- gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), data = CC_SR, knots=knots)
gam.check(cc_g, rep=100)
```

```{r}
summary(cc_g)
```
```{r}
plot(cc_g, residuals = TRUE)
```
I'll be honest, the Gaussian doesn't look too terrible. 

#### Poisson

```{r}
cc_p = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=poisson, data = CC_SR, knots=knots)
gam.check(cc_p, rep=100)
```

```{r}
plot(cc_g, residuals = TRUE, cex=5)
```
Honestly the diagnostic plots don't look much different to me, though the qqplot
is worse.

```{r}
summary(cc_p)
```
The poisson model explains less of the variance in the data. Month is no longer a significant predictor, as it was with the gaussian model.

#### Negative Binomial

```{r}
cc_nb = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=nb, data = CC_SR, knots=knots)
gam.check(cc_nb, rep=100)
```

```{r}
plot(cc_nb, residuals = TRUE, cex=5)
```

```{r}
summary(cc_nb)
```

The r2 and deviance explained are about the same as poisson

#### Tweedie
```{r}
cc_tw = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=tw, data = CC_SR, knots=knots)
gam.check(cc_tw, rep=100)
```
```{r}
plot(cc_tw, residuals = TRUE, cex=5)
```
Improved qq from neg binomial and poisson

```{r}
summary(cc_tw)
```

Slightly better rsq from nb and poisson -- month is again significant

### Kangaroo rat exclosures
#### Poisson
To speed this up, I'm jumping straight to the Poisson.

```{r}
ec_p = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=poisson, data = EC_SR, knots=knots)
gam.check(ec_p, rep=100)
```

Not terrible

```{r}
plot(ec_p, residuals = TRUE, cex=5)
```
```{r}
summary(ec_p)
```

Seasonal signal not significant?

```{r}
ec_nb = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=nb, data = EC_SR, knots=knots)
gam.check(ec_nb, rep=100)
```
I'd call this equally mediocre compared to poisson

#### Negative Binomial
```{r}
plot(ec_nb, residuals = TRUE, cex=5)
```

```{r}
summary(ec_nb)
```
A smidge more deviance explained

#### Tweedie

```{r}
ec_tw = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=tw, data = EC_SR, knots=knots)
gam.check(ec_tw, rep=100)
```
this looks good to me
```{r}
plot(ec_tw, residuals = TRUE, cex=5)
```
```{r}
summary(ec_tw)
```

Tweedie looks best.  Month is even almost significant

### Rodent Exclosure

Last but not least

#### Poisson

```{r}
xc_p = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=poisson, data = XC_SR, knots=knots)
gam.check(xc_p, rep=100)
```

```{r}
plot(xc_p, residuals = TRUE, cex=5)
```

```{r}
summary(xc_p)
```

#### Negative Binomial

```{r}
xc_nb = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=nb, data = XC_SR, knots=knots)
gam.check(xc_nb, rep=100)

```
```{r}
plot(xc_nb, residuals = TRUE, cex=5)

```
```{r}
summary(xc_nb)
```

#### Tweedie

```{r}
xc_tw = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time), family=tw, data = XC_SR, knots=knots)
gam.check(xc_tw, rep=100)
```

```{r}
plot(xc_tw, residuals = TRUE, cex=5)
```
```{r}
summary(xc_tw)
```

I feel like Tweedie looks the best for all three treatments.  Now let's add season-trend interactions... this could be either interesting or not, since seasonal signal was only marginally significant at best for species richness.


## Season-Trend interactions
Control plots:

```{r}
cc_tw_X = gam(nsp ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = CC_SR, family=tw, method = "REML", knots = knots)
summary(cc_tw_X)
```
```{r}
plot(cc_tw_X, pers=TRUE)
```
Kangaroo rat exclosures:
```{r}
ec_tw_X = gam(nsp ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = EC_SR, family=tw, method = "REML", knots = knots)
summary(ec_tw_X)
```
```{r}
plot(ec_tw_X, pers=TRUE)
```
Rodent exclosures:

```{r}
xc_tw_X = gam(nsp ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = XC_SR, family=tw, method = "REML", knots = knots)
summary(xc_tw_X)
```
```{r}
plot(xc_tw_X, pers=TRUE)
```
Testing each of the models to see if its better than additive

```{r}
cc_tw_X_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12) 
                     + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
                     data = CC_SR, method = "ML", knots = knots,family=tw)
cc_tw_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12), data = CC_SR, 
                  method = "ML", knots = knots, family = tw)
AIC(cc_tw_X_test, cc_tw_test)
```
Looks like the interaction is significant

Kangaroo rat exclosures?

```{r}
ec_tw_X_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12) 
                     + ti(Time, month, bs = c("cr","cc"), k = c(10, 12)), 
                     data = EC_SR, method = "ML", knots = knots, family = tw)
ec_tw_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12), data = EC_SR, 
                  method = "ML", knots = knots, family = tw)
AIC(ec_tw_X_test, ec_tw_test)
anova(ec_tw_test,ec_tw_X_test)
```
```{r}
xc_tw_X_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) 
                     + s(month, bs = "cc",k = 12)
                     + ti(Time, month, bs = c("cr","cc") , k = c(10, 12)), 
                     data = XC_SR, method = "ML", knots = knots, family = tw)
xc_tw_test <- gam(nsp ~ s(Time, bs = "cr", k = 10) + 
                    s(month, bs = "cc",k = 12) , data = XC_SR, 
                  method = "ML", knots = knots, family = tw)
AIC(xc_tw_X_test, xc_tw_test)
anova(xc_tw_test,xc_tw_X_test)
```

It looks like season-trend interaction is significant for controls and krat exclosures but not total exclosures.


## Checking Plot effects

Whatever the best model was coming out of the  The
random effect for plot is added as s(plot, bs="re") and that term is checked for
significance.

# Control Plots
```{r}
CC_SR$plot = as.factor(CC_SR$plot)
cc_tw_plot = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot, bs="re"), family=tw, data = CC_SR, knots=knots)
summary(cc_tw_plot)
```

No significant effect and the edf is low

```{r}
EC_SR$plot = as.factor(EC_SR$plot)
ec_tw_plot = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot,bs="re"), family=tw, data = EC_SR, knots=knots)
summary(ec_tw_plot)
```

plot might be a little bit significant

```{r}
XC_SR$plot = as.factor(XC_SR$plot)
xc_tw_plot = gam(nsp ~ s(month, bs = "cc", k = 12) + s(Time) + s(plot,bs="re"), family=tw, data = XC_SR, knots=knots)
summary(xc_tw_plot)
```
Don't need the plot random effect for any of the controls - though it is kinda
marginal for the kangaroo rat exclosures.

## PLOTS!
Control with data: I chose the gam with tweedie distribution, with month-time interaction, no plot effect
```{r}
times = select(CC_SR,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
cc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(cc_tw_X_test,  newdata = cc_pdat, type="response",se.fit = TRUE)
cc_pdat <- transform(cc_pdat, fitted=p$fit, se=p$se.fit)
cc_pdat = cc_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_cc = ggplot(aes(x = date, y = fitted), data = cc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'green') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Controls: SR ~ te(month,Time), tweedie") +
  xlab("Date") + ylab("Species richness") +
  theme_classic()
p_cc + geom_point(data=CC_SR, aes(x=date,y=nsp))
```
Kangaroo rat exclosures with data: gam with tweedie, with interaction term, no plot term
```{r}
times = select(EC_SR,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
ec_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(ec_tw_X_test,  newdata = ec_pdat, type="response",se.fit = TRUE)
ec_pdat <- transform(ec_pdat, fitted=p$fit, se=p$se.fit)
ec_pdat = ec_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_ec = ggplot(aes(x = date, y = fitted), data = ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'blue') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Kangaroo rat exclosures: SR ~ te(month,Time), tweedie") +
  xlab("Date") + ylab("Species richness") +
  theme_classic()
p_ec + geom_point(data=EC_SR, aes(x=date,y=nsp))
```
total exclosures: gam with tweedie, no interaction term, no plot term
```{r}
times = select(XC_SR,date,month,Time) %>% unique()
want <- seq(1, nrow(times), length.out = 50)
xc_pdat <- with(times, data.frame(Time = Time[want], date = date[want], 
                                 month = month[want]))
p  <- predict(xc_tw,  newdata = xc_pdat, type="response",se.fit = TRUE)
xc_pdat <- transform(xc_pdat, fitted=p$fit, se=p$se.fit)
xc_pdat = xc_pdat %>% mutate(lower = fitted - 1.96*se, upper = fitted + 1.96*se)
transition = as.Date("2015-03-15", format="%Y-%m-%d")
p_xc = ggplot(aes(x = date, y = fitted), data = xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') +
  geom_line(color = 'red') +
  geom_vline(xintercept =  as.numeric(transition)) +
  ggtitle("Rodent Exclosures: SR ~ te(month,Time), tweedie") +
  xlab("Date") + ylab("Species Richness") +
  theme_classic()
p_xc + geom_point(data=XC_SR, aes(x=date,y=nsp))
```


```{r}
transition = as.Date("2015-03-15", format="%Y-%m-%d")
ggplot(aes(x=date, y=fitted), data=xc_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = 'gray90') + 
  geom_line(color = 'red') +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = ec_pdat, fill = 'gray90') +
  geom_line(color = 'blue', data=ec_pdat) +
  geom_ribbon(aes(ymin = lower, ymax = upper), data = cc_pdat, fill = 'gray90') +
  geom_line(color = 'green', data=cc_pdat) +
  geom_vline(xintercept =  as.numeric(as.Date('2015-03-20'))) +
  ggtitle("All treatments: SR ~ te(month, Time), tweedie") +
  xlab("Date") + ylab("Species Richness") +
  theme_classic()
```


### Comparing two surfaces:

I'm trying to do this following Wood (Section 7.2.5 in his book). 

I'm going to combine the XC and EC datasets into 1 data frame and add some
dummy variables to identify which treatment each comes from.

```{r}
EC_XC=c()
EC_XC = rbind(XC_SR, EC_SR)
EC_XC$XC = c(rep(1,nrow(XC_SR)), rep(0,nrow(XC_SR)))
EC_XC$EC = 1- EC_XC$XC
```

Ok, now here's his approach of fitting a common surface and difference
surface:

```{r}
same = gam(nsp ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)),            
               data = EC_XC, family=tw, method = "REML", knots = knots)
diff = gam(nsp ~ te(Time,month, bs = c("cr","cc"), k =c(10,12)) +
             te(Time,month, by=XC, bs = c("cr","cc"), k =c(10,12)),            
               data = EC_XC, family=tw, method = "REML", knots = knots)
AIC(same,diff)
```
```{r}
anova(diff)
```
```{r}
summary(diff)
```
```{r}
plot(diff, pers=TRUE)
```

This result seems more robust than the Dipo result. Both AIC and ANOVA point in the same direction, that modelling the difference between the krat and exclosure surfaces
improves the model. The surface model of diff suggests that the difference is
high at first (under the old treatments), and then they converge and diverge. One thing to think about with this type of approach is that the differences before the flip might 
create the significant diff effect. This analysis may be better conducted
on just the dat since the flip.



 
```{r}

```
