---
title: "GAM_nongaussian"
output: html_notebook
---

```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)

source('gam_functions.R')
source('data_functions.R')
```

# Portal Plot Flip using gam()

This notebook tries to apply factor-smooth approach outlined by Gavin 
Simpson in his [blog post](https://www.fromthebottomoftheheap.net/2017/10/10/difference-splines-i/)
and combine it with the trend-season appraoch outlined in his other series of
posts: [modelling seasonal data with GAMS](https://www.fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/),
[are some seasons warming more than others?](https://www.fromthebottomoftheheap.net/2015/11/23/are-some-seasons-warming-more-than-others/)

Here we go:

## Load Data

Data is coming straight from the repo. This code is extracting the dipo 
abundances from the straight controls (CC), and the opened krat (EC) and rodent 
(XC) exclosures. This data runs from period 415 (3/13/2013) to present

```{r}
fulldata = get_data()
# get average value by treatment
data = aggregate(fulldata$abundance, by=list(censusdate = fulldata$censusdate, 
                                             species = fulldata$species,
                                             treatment = fulldata$treatment,
                                             numericdate = fulldata$numericdate),FUN=mean)

dipos = dplyr::filter(data,species %in% c('DO','DM','DS'))
dipoN = aggregate(dipos$x,by=list(censusdate=dipos$censusdate,
                                          treatment=dipos$treatment,
                                          numericdate=dipos$numericdate),FUN=sum)
dipoN=plyr::rename(dipoN,c('x'='abundance'))
dipoN$month = as.numeric(format.Date(dipoN$censusdate, "%m"))

write.csv(dipoN,'Dipo_abundance_by_treatment.csv',row.names = F)

dipoN = read.csv('Dipo_abundance_by_treatment.csv')
dipoN$censusdate = as.Date(dipoN$censusdate)
dipoN
#knots <- list(month = c(0.5, seq(1, 12, length = 10), 12.5))
```


## Basic Season-trend model with factors

Here's a grap hof the basic patterns in the data by treatment type. We see
the pure control plots (green) are relative stable through time (dip in most 
recent months) and the increase of the kangaroo rats on the two opened plots, 
with the number of kangaroo rats appearing higher on the opened rodent 
exclosures (blue) than the kangaroo rat exclosures (red)


```{r}
ggplot(dipoN, aes(x = censusdate, y = abundance, colour = treatment)) +
    geom_point() +
    geom_smooth(method = 'loess', se = FALSE) +
    scale_colour_brewer(type = 'qual', palette = 'Dark2') +
    theme(legend.position = 'top')
```

## GAM model

This is strictly following Gavin's post

```{r pressure, echo=FALSE}
# add site code to the model as a factor
m <- gam(abundance ~ treatment + s(numericdate, by = treatment), data = dipoN)
summary(m)
plot(m, shade = TRUE, pages = 1, scale = 0)
```


 This model has fit a smooth to each of the lakes (sitecode) independently.  We want to compare them. 
 To compare the smooths pair-wise we need to get the xp matrix, used for "prediction"
 
```{r pressure, echo=FALSE}
# add site code to the model as a factor
tmin = min(dipoN$numericdate)
tmax = max(dipoN$numericdate)
pdat <- expand.grid(numericdate = seq(tmin, tmax, length = 200),
                    treatment = c('CC', 'EC', 'XC'))
xp <- predict(m, newdata = pdat, type = 'lpmatrix')

# function to compare treatments pair-wise
smooth_diff <- function(model, newdata, f1, f2, var, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}
```

```{r, echo=False}
# use the function to compare all pairs of treatments
comp1 <- smooth_diff(m, pdat, 'CC', 'EC', 'treatment')
comp2 <- smooth_diff(m, pdat, 'CC', 'XC', 'treatment')
comp3 <- smooth_diff(m, pdat, 'EC', 'XC', 'treatment')
comp <- cbind(date = seq(tmin, tmax, length = 200),
              rbind(comp1, comp2, comp3))
ggplot(comp, aes(x = date, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 2) +
    coord_cartesian(ylim = c(-8,8)) +
    labs(x = NULL, y = 'Difference in # of dipos')
```
