---
title: "Dipo_GAM_analysis"
author: EMC 1/2018
output: html_notebook
---



```{r setup, include=FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)


```

# Portal Plot Flip using gam()

This project is exploring the effect of a resident (animal) community on the 
ability of an additional species to colonize a patch. Things like priority effects 
imply that it should be difficult for a new species to move in to an occupied patch, 
but coexistence theories generally predict that the ability of a community of species 
to coexist and the dominance heirarchy of that community should be independent of who 
got there first.

Main question: does the presence of an established community on a patch slow or prevent 
the "invasion" of a species that should be dominant in the area?

At the Portal Project we have the perfect experimental setup to test this. Kangaroo rats
(genus Dipodomys, of which we have 3 species at Portal) are the clear dominant members of
the rodent community. The Portal Project consists of 24 fenced plots divided into 3 
treatment types: controls (large gates in fences, all rodents have access), kangaroo 
rat exclosures (small gates in fences, kangaroo rats are excluded but all other species 
have access), and total removals (no gates in fences, no rodents have access). 

In March 2015 we changed the treatment on a subset of the plots, cutting new gates or 
enlarging existing gates on restricted plots to allow access to all rodents (including 
kangaroo rats). For this project we're comparing plots that used to be k-rat exclosures 
and became controls (I label these treatment 'EC'), plots that used to be total rodent 
removals and became controls ('XC'), and plots that were always controls and remained 
controls ('CC'). So k-rats are invading an established community in the 'EC' treatment, 
they (along with the other species) are invading an empty patch in the 'XC' treatment, 
and the "background" control community is represented in the 'CC' treatment. 

We're using GAMs to ask whether the number of kangaroo rats on these treatment types 
remains different over time or whether they all converve to the 'CC' eventually. If the
treatments do converge, we want to know at what point that happens. We combined all 3 
species of dipodomys in this calculation, because all 3 species were excluded from the 
k-rat exclosure plots.


## Load Data
Data contains columns for plot, censusdate, numericdate (so we don't have to worry about date 
formatting), treatment, and dipos. 


```{r}

dipoN = read.csv('Dipo_counts.csv')

dipoN$censusdate = as.Date(dipoN$censusdate)
#dipoN

```


## Basic plot of data

Here's a graph of the data by treatment type. We see the long-term control plots 
(CC, green) are relatively stable through time (dip in most recent months) and the 
increase of the kangaroo rats on the two opened plots after the treatment change, 
with the number of kangaroo rats appearing higher on the opened rodent exclosures 
(XC, purple) than the kangaroo rat exclosures (EC, red). Vertical line is when we 
did the treatment change. (before the treatment change, both EC and XC treatments 
should have zero k-rats. A few show up in the data because sometimes they make it 
into the fences anyway, but they are promptly removed.)


```{r}
# ggplot(dipoN, aes(x = censusdate, y =dipos, colour = treatment)) +
#     geom_point() +
#     geom_smooth(method = 'loess', se = TRUE) +
#     scale_colour_brewer(type = 'qual', palette = 'Dark2') +
#     theme(legend.position = 'top') +
#   geom_vline(xintercept=as.Date('2015-04-10'))
```

## GAM model - ordered factor for treatment type


```{r, echo=FALSE}
# create ordered factors for treatment and plot
dipoN <- mutate(dipoN,
                 oTreatment = ordered(treatment, levels = c('CC','EC','XC')),
                 oPlot      = ordered(plot),
                 plot       = factor(plot))

# The model: includes plot-specific smooth differences
# need `select` here as there is a little issue with identifiability as plot is nested
# in treatment

m2 <- gam(dipos ~ oPlot + oTreatment + s(numericdate, k = 20) +
              s(numericdate, by = oTreatment, k = 15) +
              s(numericdate, by = oPlot),
          data = dipoN, method = 'REML', family = poisson, select = TRUE)



plot(m2, shade = TRUE, pages = 1,  scale=0,seWithMean = T)

```



 
```{r, echo=FALSE}
# is this a good model?
summary(m2)
gam.check(m2, rep=200)

```
```{r, echo=F}
# Look at the treatment effect smooths on count scale. Requires us to exclude the
# plot-level effects.

# Data to predict at [plot=4 is a dummy plot, so the data frame variables match the model]
np <- 500
treatEff <- with(dipoN,
                 expand.grid(censusdate = seq(min(censusdate), max(censusdate), length = np),
                             treatment  = c('CC','EC','XC'),
                             plot       = 4))

# create derived variables from the data we want to predict at
treatEff <- transform(treatEff,
                      oPlot       = ordered(plot),
                      oTreatment  = ordered(treatment, levels = c('CC','EC','XC')),
                      numericdate = as.numeric(censusdate) / 1000)

# terms to exclude; must be named exactly as printed in `summary(model)` output
exVars <- c('oPlot', paste0('s(numericdate):oPlot', c(5,6,7,11,13,14,17,18,24)))

# actually predict, on link scale so we can get proper CIs, exclude
treatPred <- as.data.frame(predict(m2, treatEff, type = 'link', se.fit = TRUE,
                                   exclude = exVars))

# bind predictions to data we predicted at
treatPred <- cbind(treatEff, treatPred)

# extract inverse of link function from model
ilink <- family(m2)$linkinv

# form 95% bayesian credible interval / frequentist across-function confidence interval
treatPred <- transform(treatPred, Fitted = ilink(fit),
                       Upper = ilink(fit + (2 * se.fit)),
                       Lower = ilink(fit - (2 * se.fit)))

# plot
p.plt <- ggplot(treatPred, aes(x = censusdate, y = Fitted)) +
    geom_point(data = dipoN, mapping = aes(y = dipos, colour = treatment)) +
    geom_ribbon(aes(ymax = Upper, ymin = Lower, fill = treatment),
                alpha = 0.2) +
    geom_line(aes(colour = treatment)) +
    labs(y = 'Count', x = NULL) +
    theme(legend.position = 'top') +
    scale_colour_brewer(name = 'Treatment', type = 'qual', palette = 'Dark2') +
    scale_fill_brewer(name = 'Treatment', type = 'qual', palette = 'Dark2') +
  geom_vline(xintercept=as.Date('2015-04-10'))
p.plt

```
```{r, echo=F}
# plot treatment trends
exVars2 <- c(exVars, "oTreatment")
treatPred2 <- predict(m2, treatEff, type = 'terms', exclude = exVars2)
treatPred2 <- rowSums(treatPred2)
treatPred2 <- cbind(treatEff, Fitted = treatPred2)
## plot
p.plt2 <- ggplot(treatPred2, aes(x = censusdate, y = Fitted)) +
    geom_line(aes(colour = treatment)) +
    labs(y = 'Count', x = NULL) +
    theme(legend.position = 'top') +
    scale_colour_brewer(name = 'Treatment', type = 'qual', palette = 'Dark2')
p.plt2
```


```{r, echo=F}
## Compute pairwise differences of smooths when fitted using ordered factors

osmooth_diff <- function(model, newdata, smooth_var, f1, f2, var, alpha = 0.05,
                         unconditional = FALSE, ...) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix', ...)
    ## reference level
    ref_level <- levels(newdata[[var]])[1L]
    ref_smooth <- grepl(paste0("s\\(", smooth_var, "\\)\\.{1}[[:digit:]]+$"), colnames(xp))
    c1 <- ref_smooth | grepl(f1, colnames(xp))
    c2 <- ref_smooth | grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

## Compute pairwise diffs
d1 <- osmooth_diff(m2, treatEff, "numericdate", "CC", "EC", var = "oTreatment")
d2 <- osmooth_diff(m2, treatEff, "numericdate", "CC", "XC", var = "oTreatment")
d3 <- osmooth_diff(m2, treatEff, "numericdate", "EC", "XC", var = "oTreatment")

diffs <- rbind(d1, d2, d3)
diffs <- with(treatEff, cbind(censusdate, diffs)) # bind on censusdate for plotting

## plot difference, facetted
ggplot(diffs, aes(x = censusdate, y = diff)) +
    geom_ribbon(aes(ymax = upper, ymin = lower), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 1) +
    labs(y = 'Difference', x = NULL)

```

