---
title: "Seasonal GAMs with Portal"
author: "Morgan Ernest"
output: html_notebook
---
This is my attempt to implement Gavin Simpson's GAM tutorial with timeseries
data with Portal Rodent Abundance.

## Season and Trend - non interactive

First step: Load and process the data. For the seasonal analyses, Gavin used
a continuous time variable for the trend component. He did this by taking the 
data object and converting it to its numeric basis and then dividing by 1000
to make it manageable. Season is with within year Julian Day.

the notation 'cet' is just a holdover from Gavin's code because I was lazy.

Don't forget to set your working directory!
```{r Load Data}
require("mgcv")
data = read.csv("raw_abundance.csv")
data$date = as.Date(data$date, format="%m/%d/%Y")
cet <- transform(data, Time = as.numeric(date) / 1000)
cet = transform(cet, Season = as.numeric(format(date, "%j")))
```

What does a seasonal GAM even look like? This model is fitting a rodent abundance
as a function of a smooth of Season and a smooth of Time independently. The
season smooth is set to a cyclic spline so that January and December are similar.
12 knots are specified for the 12 months. In a later tutorial, he decides that
for seasonal fits, we're better off fixing the locations of the knots so that
January and December aren't just assumed to be the same. Turns out that because season is implementd continuously, its a little more complicated in this model to do that, but we will do it for the interaction stuff later.

The response distribution is just the default, Gaussian (?), but is probably incorrect since this is count data not temperature data like Gavin was using. 
But let's not complicate things more than they already are at the moment.

```{r gam fit}
m <- gamm(abundance ~ s(Season, bs = "cc", k = 12) + s(Time), data = cet)
summary(m$gam)
```
Both season and trend are significant. An r2 of .363 is pretty damn good for
Portal, to be honest. 

```{r gam plots}
layout(matrix(1:2, ncol = 2))
plot(m$gam, scale = 0)
layout(1)
```

Looks like a model distribution of abundance, highest at midyear and a generally
increasing trend in abundance, with lots of wiggles.

```{r autocorrelation plots}
layout(matrix(1:2, ncol = 2))
acf(resid(m$lme), lag.max = 36, main = "ACF")
pacf(resid(m$lme), lag.max = 36, main = "pACF")
layout(1)
```

Ah, autocorrelation. Not suprising. Looks like there is significant autocorrelation upto about lab 7 or 8. pacf suggests that could be echos from the first lag(?).

Let's add within year autocorrelation! Here is Gavin's code for fitting a bunch
of AR models in an automated fashion. Since there were signals up to 8, we'll
look up to p=8.

```{r add AR residuals}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)
for (i in 1:8) {
  m <- gamm(abundance ~ s(Season, bs = "cc", k = 12) + s(Time),
            data = cet, method = "REML", control = ctrl,
            correlation = corARMA(form = ~ 1 | yr, p = i))
  assign(paste0("m", i), m) 
}
anova(m$lme, m1$lme, m2$lme, m3$lme, m4$lme, m5$lme, m6$lme, m7$lme, m8$lme)
```

We might be able to argue that we don't need an AR process modelled, but
given the ACF results, I think its safer to go with the AR(2) model. Let's
look at that:

```{r}
layout(matrix(1:2, ncol = 2))
plot(m2$gam, scale = 0)
layout(1)
```
Our seasonal response is more bimodal and our trend is now linear.

```{r}
layout(matrix(1:2, ncol = 2))
res <- resid(m2$lme, type = "normalized")
acf(res, lag.max = 36, main = "ACF - AR(2) errors")
pacf(res, lag.max = 36, main = "pACF- AR(2) errors")
layout(1)
```
We have some indication of between year autocorrelation, at about lag 12, but
I'm going to pull a Simon Wood here and say "Let's not worry about that now"

I'd have to go back to Gavin's post to see what this graph is really about, but
I suspect its showing how the AR() impacts your trend estimates to demonstrate
why this is potentially important.
```{r}
want <- seq(1, nrow(cet), length.out = 39)
pdat <- with(cet, data.frame(Time = Time[want], date = date[want], 
                             Season = Season[want]))

## predict trend contributions
p  <- predict(m$gam,  newdata = pdat, type = "terms", se.fit = TRUE)
p1 <- predict(m1$gam, newdata = pdat, type = "terms", se.fit = TRUE)
p2 <- predict(m2$gam, newdata = pdat, type = "terms", se.fit = TRUE)
p3 <- predict(m3$gam, newdata = pdat, type = "terms", se.fit = TRUE)

pdat <- transform(pdat, p  = p$fit[,2],  se  = p$se.fit[,2], p1 = p1$fit[,2], se1 = p1$se.fit[,2],
                  p2 = p2$fit[,2], se2 = p2$se.fit[,2], p3 = p3$fit[,2], se3 = p3$se.fit[,2])

op <- par(mar = c(5,4,2,2) + 0.1)
ylim <- with(pdat, range(p, p1, p2, p3))
ylim[1] <- floor(ylim[1])
ylim[2] <- ceiling(ylim[2])
ylab <- expression(Abundance ~ (Individuals/ha ~ centred))
plot(abundance - mean(abundance) ~ date, data = cet, type = "n", 
     ylab = ylab, ylim = ylim)
lines(p  ~ date, data = pdat, col = "black")
lines(p1 ~ date, data = pdat, col = "red")
lines(p2 ~ date, data = pdat, col = "blue")
lines(p3 ~ date, data = pdat, col = "forestgreen", lwd = 1)
legend("topleft", legend = c("Uncorrelated Errors", paste0("AR(", 1:3, ") Errors")),
       bty = "n", col = c("black","red","blue","forestgreen"), lty = 1, lwd = c(1,1,1))
par(op)
```

Back to the seasonal model: Like the other modelling examples (when I worked 
through earlier tutorials in this series) this one also says that the increase 
in Abundance is linear across years. It does seem to pick up a bimodal seasonal dynamic in abundance. Is this true or a signal of shifting seasonal changes?
I think our next tutorial will clarify this.

## Interacting Season and Trend

So, does the seasonal signal change through time at Portal? Here's the
attempt to use Gavin's blog post on the Portal data. Again, the response
distribution is the default which is probably not correct.

This time instead of continuous time, Gavin has reverted to Year and Month. I
have done the same. As I experienced earlier when I tried to implement the
new knot placement for season in the previous model, it is easier to do with
month as a discrete value.

I'm keeping Gavin's knot structure specified for the yr term. I suspect this
helps with model comparison later?


```{r}
knots <- list(mo = c(0.5, seq(1, 12, length = 10), 12.5))
m0 <- gamm(abundance ~ te(yr, mo, bs = c("cr","cc"), k = c(10,12)),
           data = cet, method = "REML", knots = knots)
```
The first model is without any AR term for the residuals.

```{r}
plot(acf(resid(m0$lme, type = "normalized")))
```

This shows a similar autocorrelation pattern as earlier. We'll do the same
search for the best AR order.

```{r}
ctrl <- list(niterEM = 0, optimMethod="L-BFGS-B", maxIter = 100, msMaxIter = 100)

for (i in 1:8) {
  m <- gamm(abundance ~ te(yr, mo, bs = c("cr","cc"), k = c(10,12)),
            data = cet, method = "REML", control = ctrl, knots = knots,
            correlation = corARMA(form = ~ 1 | yr, p = i))
  assign(paste0("m", i), m) 
}
anova(m1$lme, m2$lme, m3$lme, m4$lme, m5$lme, m6$lme, m7$lme, m8$lme)
```

Like last time, the AR2 model seems to be best.

```{r}
plot(acf(resid(m2$lme, type = "normalized")))

```
This looks really good! So what does the surface of the interaction look like?
```{r}
plot(m2$gam, pers = TRUE)
```
Does seem to be a strong change in seasonality with time!

Let's do what Gavin did and see how the seasonal abundance estimates differ
in 1978 and 2014.

```{r}
pdat <- with(cet, data.frame(yr = rep(c(1978, 2014), 
                                        each = 100),
                             mo = rep(seq(1, 12, 
                                              length = 100),
                                          times = 2)))

pred <- predict(m2$gam, newdata = pdat, se.fit = TRUE)
crit <- qt(0.975, df = df.residual(m$gam)) # ~95% interval critical t
pdat <- transform(pdat, fitted = pred$fit, se = pred$se.fit, fYear = as.factor(yr))
pdat <- transform(pdat,
                  upper = fitted + (crit * se),
                  lower = fitted - (crit * se))

library(ggplot2)

p1 <- ggplot(pdat, aes(x = mo, y = fitted, group = fYear)) +
  geom_ribbon(mapping = aes(ymin = lower, ymax = upper,
                            fill = fYear), alpha = 0.2) + # confidence band
  geom_line(aes(colour = fYear)) +    # predicted temperatures
  theme_bw() +                        # minimal theme
  theme(legend.position = "top") +    # push legend to the top
  labs(y = expression(abundance), x = NULL) +
  scale_fill_discrete(name = "Year") + # correct legend name
  scale_colour_discrete(name = "Year") +
  scale_x_continuous(breaks = 1:12,   # tweak where the x-axis ticks are
                     labels = month.abb, # & with what labels
                     minor_breaks = NULL)
p1
```

Our model suggests that there was little seasonality in the data early on, and
whatever seasonality there was was in caused by higher abundance in winter. Now,
we have a different pattern, strong increased abundance in summer.

We can plot month trends:

```{r}
pdat2 <- with(cet,
              data.frame(yr = rep(1978:2014, each = 12),
                         mo = rep(1:12, times = 37)))

pred2 <- predict(m$gam, newdata = pdat2, se.fit = TRUE)

pdat2 <- transform(pdat2,
                   fitted = pred2$fit,  # predicted values
                   se = pred2$se.fit,   # standard errors
                   fMonth = factor(month.abb[mo], # month as a factor
                                   levels = month.abb))
pdat2 <- transform(pdat2,
                   upper = fitted + (crit * se), # upper and...
                   lower = fitted - (crit * se)) # lower confidence bounds


p2 <- ggplot(pdat2, aes(x = yr, y = fitted, group = fMonth)) +
  geom_line(aes(colour = fMonth)) +   # draw trend lines
  theme_bw() +                        # minimal theme
  theme(legend.position = "none") +   # no legend
  labs(y = expression(Abundance), x = NULL) +
  facet_wrap(~ fMonth, ncol = 6) +    # facet on month
  scale_y_continuous(breaks = seq(0, 100, by = 10),
                     minor_breaks = NULL) # nicer ticks
p2
```

Every month has seen an increase in abundance, but the strongest increases are
in the summer.

We can group things by season, because why not?

```{r}
pdat2$Quarter <- NA
pdat2$Quarter[pdat2$mo %in% c(12,1,2)] <- "Winter"
pdat2$Quarter[pdat2$mo %in% 3:5] <- "Spring"
pdat2$Quarter[pdat2$mo %in% 6:8] <- "Summer"
pdat2$Quarter[pdat2$mo %in% 9:11] <- "Autumn"
pdat2 <- transform(pdat2,
                   Quarter = factor(Quarter,
                                    levels = c("Spring","Summer","Autumn","Winter")))

p3 <- ggplot(pdat2, aes(x = yr, y = fitted, group = fMonth)) +
  geom_line(aes(colour = fMonth)) +   # draw trend lines
  theme_bw() +                        # minimal theme
  theme(legend.position = "top") +    # legend on top
  scale_fill_discrete(name = "Month") + # nicer legend title
  scale_colour_discrete(name = "Month") +
  labs(y = expression(abundance), x = NULL) +
  facet_grid(Quarter ~ ., scales = "free_y") # facet by Quarter
p3
```

So, is the interaction actually a better fit than the non-interactive model?

Here I try to implement Gavin's approach with the Portal data. To compare
the interactive with the non-interactive model, we need to rewrite them
so that one is a nested versionof the other: 

y ~ s(x1) + s(x2)
y ~ s(x1) + s(x2) + ti(x1, x2)

Where season and trend are fit separately in both version and with the only
difference being the addition of an interaction. To make sure that the model
are otherwise identical, we have to pull the AR fit from the interaction
model and force it to be used by both models.

```{r}
phi <- unname(intervals(m2$lme, which = "var-cov")$corStruct[, 2])
m1 <- gamm(abundance ~ s(yr, bs = "cr", k = 10) + s(mo, bs = "cc", k = 12) +
               ti(yr, mo, bs = c("cr","cc"), k = c(10, 12)),
           data = cet, method = "ML", control = ctrl, knots = knots,
           correlation = corARMA(value = phi, fixed = TRUE, form = ~ 1 | yr, p = 2))
m0 <- gamm(abundance ~ s(yr, bs = "cr", k = 10) + s(mo, bs = "cc", k = 12),
           data = cet, method = "ML", control = ctrl, knots = knots,
           correlation = corARMA(value = phi, fixed = TRUE, form = ~ 1 | yr, p = 2))
anova(m0$lme, m1$lme)

```

Strong season x year interaction! Now to figure out a) if the response distribution is valid and how to implement with a different data model.

```{r}
summary(m1$gam)
```
```{r}
summary(m1$lme)
```

